### 2013-07-24 writeups/context-dependence.tex
.

If the approximation described above holds,
then we can write the probability that we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$
using the pattern frequencies at the root.
Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{m+2\ell}(0) = \rho$ be the (long) pattern seen there at the root,
and write $\pi(z)$ for the frequency of pattern $z$ in $X(0)$, i.e.\ $\P\{\rho=z\}=\pi(z)$.
% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
% \[
%     \P\{ \rho=z \mid X_{I-\ell}^{m+2\ell}(t_X)=x\} \propto \pi(z) p_{m+2\ell}(t,z,x) .
% \]
The probability that we see $x$ and $y$ at the random location $I$ is
\begin{align} \label{eqn:phylo_likelihood}
    \P\{X_{I-\ell}^{m+2\ell}(t_X)=x \text{ and } Y_I^\ell(t_Y)=y \} = \sum_{z \in \calS^{m+2\ell}} \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) .
\end{align}
This can be computed without much more effort than the simpler case above,
given the frequencies at the root, as described in greater generality below.

%%%%
\subsection{The root distribution}

@@

%%%%%
\subsubsection{Mixing}

There can be significant confounding of the frequencies at the root with the other parameters,
which results in poor mixing.
(See CpG results, below.)
To see why this is, consider the following back-of-the-envelope calculation.
Take $x \neq y$ to be single bases (so, set $\ell=0$ and $w=1$),
so that $N(x,y)$ gives the number of times an $x$ and a $y$ are homologous.
Let $q(x,y)$ be the instantaneous rate that $x \to y$ (averaged somehow over context?),
so that $N(x,y)$ should be, roughly,
\begin{align*}
  \sum_z \pi(z) q(z,x) q(z,y) &\approx t \pi(x) (1+q(x,x)) q(x,y) + t \pi(y) q(y,x) (1+q(y,y)) ,  
\end{align*}
where as usual $q(x,x) = (-1) \sum_{z \neq x} q(x,z)$ is (minus) the total jump rate out of $x$.

So, when we propose a change to $\pi$ we might want to also adjust the rates accordingly to keep this more-or-less constant:
if we chnage $\pi(x) \mapsto \pi(x) + \epsilon_x$, we want $\delta_{xy}$ to satisfy
\begin{align*}
  \sum_z \pi(z) q(z,x) q(z,y) = \sum_z (\pi(z) + \epsilon_z) ( q(z,x) + \delta_{zx} ) ( q(z,y) + \delta_{zy} ) ,
\end{align*}
or, to first order in $\delta$,
\begin{align*}
  0 &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z \pi(z) \delta_{zx} q(z,y) + \sum_z \pi(z) q(z,x) \delta_{zy} \\
   &\approx \sum_z \epsilon_z q(z,x) q(z,y) + \sum_z (\pi(z)+\epsilon_z) \delta_{zx} q(z,y) + \sum_z (\pi(z)+\epsilon_z) q(z,x) \delta_{zy} \\
   &\approx \epsilon_x (1+q(x,x)) q(x,y) + \epsilon_y q(y,x) (1+q(y,y)) + (\pi(y)+\epsilon_y) \delta_{yx} (1+q(y,y)) + (\pi(x)+\epsilon_x) (1+q(x,x)) \delta_{xy} \\
   &= \left( \epsilon_x q(x,y) + (\pi(x)+\epsilon_x) \delta_{xy} \right)(1+q(x,x)) + \left( \epsilon_y q(y,x) + (\pi(y)+\epsilon_y) \delta_{yx} \right)(1+q(y,y)) 
\end{align*}
One solution to this is to let $\delta_{xy} = - \epsilon_x q(x,y) / \pi(x)$.

Suppose each change is a single-base change, so that changes $\delta \mu_i$ in $\mu$ contribute to $q$ by
$\delta_{xy} = \sum_{i: x \to y} \delta \mu_i$.
Given $\delta_{xy}$, this gives a linear equation for $\delta \mu_i$ that may or may not be solvable,
but certainly is if we have all the single-base transitions as paramters.

@@

We can do similar precomputations to help with the case of selective differences:
XXX describe

@@

More generally\dots

\begin{center}
\includegraphics{peeling-schematic}
\end{center}

To compute the likelihood on a more complicated tree, consider the following,
with the three-taxon tree above as an example,
where we are counting occurrences of (long) $m+2\ell$-tuples in taxon $X$,
and (shorter) $m$-tuples in taxa $Y$ and $Z$.

We can compute the likelihood in a similar way, summing over the root state (the summation over $u$)
and the interior node ($v$):
\begin{align} \label{eqn:three_taxa_likelihood}
  \begin{split}
    & \P\{X_{I-\ell}^{m+2\ell}=x \text{ and } Y_I^\ell=y \text{ and } Z_I^\ell=z \} \\
    &\qquad = \sum_{u \in \calS^{m+2\ell}} \pi(u) p_{m,\ell}(t_C,u,z) \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{AB},u,v) p_{m+2\ell}(t_{A},v,x) p_{m,\ell}(t_B,v,y) 
  \end{split}
\end{align}
The steps for computing this are then:
\begin{enumerate}

  \item Compute $M_1(u,z) = p_{m,\ell}(t_C,u,z)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)

  \item Compute $M_2(v,z) = \sum_{v \in \calS^{m+2\ell}}  p_{m+2\ell}(t_{AB},u,v) \pi(u) M_1(u,z)$ (still a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)

  \item Compute $M_3(v,y) = p_{m,\ell}(t_B,v,y)$ (a $|\calS|^{m+2\ell} \times |\calS|^m$ matrix)

  \item Compute $M_4(x,(y,z)) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,z) M_4(v,y)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{2m}$ matrix).

\end{enumerate}
Then $M_4$ gives all the likelihoods.
This is depicted in the figure, writing $\left(e^{tG}\right)_{ab}$ for the matrix $p_{m+2\ell}(t,a,b)$,
and $P$ for the projection matrix that gives $p_{m,\ell}(t,a,b) = \left( e^{tG} P\right)_{ab}$.

In total there are $|\calS|^{3m+2\ell}$ possible data combinations.  
However, we can reduce dimensionality by replacing the last step with
\begin{enumerate}

  \item[4'.] Compute $M_4(x,w) = \sum_{v \in \calS^{m+2\ell}} p_{m+2\ell}(t_{A},v,x) M_3(v,w) M_4(v,w)$ (stored as a $|\calS|^{m+2\ell} \times |\calS|^{m}$ matrix).

\end{enumerate}
and only computing the likelihood for cases with $y=z=w$.
On a three-taxon tree it is not clear that this is desireable, since missing out on double transitions 
might be a significant loss of information,
but on larger trees it seems likely that we'd want to restrict to combinations with most sites agreeing across the tree.
(Note, however, that one should not first look at the data, observe which combinations $(y,z)$ were most common, and then restrict to only those!)

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{tasep/all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates for six different data sets of the parameter in TASEP (mutation rate multiplied by time).  
  Vertical line shows the true value.
  }
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics{writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
  \end{center}
  \caption{ 
  MCMC traces of the parameters in the Ising model across 100,000 iterations.
  The units of $\lambda$ are in mean number of changes per site.
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth,page=1]{ising/all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates of the time-scaled mutation rate parameter $\lambda t$,
  in ten different simulated datasets under the Ising model.
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth,page=2]{ising/all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates of the inverse temperature $\beta$ in the Ising model. 
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth,page=3]{ising/all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates of the magnetization parameter $\beta$ in the Ising model. 
  Vertical line shows the true value.
  }
\end{figure}

@@

%%%%%%%%
\section{Power sims}

Assess power and accuracy in the Ising model, by running replicates at these parameters:

\begin{tabular}{|rlllrrr|}
  \hline
  seqlen   &  $\lambda t$   &  $\beta$  &  $\gamma$  &  $w$ &  $\ell$   & \# reps \\
  \hline
  $10^4$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
  $10^3$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
  $10^2$    &   .2  &   1   &   .5  &   3   &   3   &   10  \\
  \hline
  $10^4$    &   .4  &   1   &   .5  &   3   &   3   &   10  \\
  $10^4$    &   .1  &   1   &   .5  &   3   &   3   &   10  \\
  $10^4$    &   .05  &   1   &   .5  &   3   &   3   &   10  \\
  \hline
  $10^4$    &   .2  &   1   &   0  &   3   &   3   &   10  \\
  \hline
  $10^4$    &   .2  &   1   &   .5  &   3   &   2   &   10  \\
  $10^4$    &   .2  &   1   &   .5  &   3   &   1   &   10  \\
  $10^4$    &   .2  &   1   &   .5  &   3   &   0   &   10  \\
  \hline
\end{tabular}

@@

  \item See if counting in nonoverlapping windows fixes the peakiness problem with TASEP. \emph{Yes, it does.}

>>>>

If the approximation described above holds,
then we can write the probability that we see (long) pattern $x$ in $X(t_X)$ juxtaposed with (short) pattern $y$ in $Y(t_Y)$
using the pattern frequencies at the root.
Concretely: pick a random location $I$ in the sequence and let $X_{I-\ell}^{m+2\ell}(0) = \rho$ be the (long) pattern seen there at the root,
and write $\pi(z)$ for the frequency of pattern $z$ in $X(0)$, i.e.\ $\P\{\rho=z\}=\pi(z)$.
% Given the corresponding pattern in $X(t_X)$, the distribution of $\rho$ is proportional to
% \[
%     \P\{ \rho=z \mid X_{I-\ell}^{m+2\ell}(t_X)=x\} \propto \pi(z) p_{m+2\ell}(t,z,x) .
% \]
The probability that we see $x$ and $y$ at the random location $I$ is
\begin{align} \label{eqn:phylo_likelihood}
    \P\{X_{I-\ell}^{m+2\ell}(t_X)=x \text{ and } Y_I^\ell(t_Y)=y \} = \sum_{z \in \calS^{m+2\ell}} \pi(z) p_{m+2\ell}(t_X,z,x) p_{m,\ell}(t_Y,z,y) .
\end{align}
This can be computed without much more effort than the simpler case above,
but requires the frequencies at the root.

\subsection{The root distribution}

@@

@@

@@

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{../tasep/all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates for six different data sets of the parameter in TASEP (mutation rate multiplied by time).  
  Vertical line shows the true value.
  }
\end{figure}

@@

\begin{figure}
  \begin{center}
    \includegraphics{../writeup-plots/selsims-2013-05-28-17-12-0275615-traces}
  \end{center}
  \caption{ 
  MCMC traces of the parameters in the Ising model across 100,000 iterations.
  The units of $\lambda$ are in mean number of changes per site.
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth,page=1]{../ising/all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates of the time-scaled mutation rate parameter $\lambda t$,
  in ten different simulated datasets under the Ising model.
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth,page=2]{../ising/all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates of the inverse temperature $\beta$ in the Ising model. 
  Vertical line shows the true value.
  }
\end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth,page=3]{../ising/all-mcmc-runs}
  \end{center}
  \caption{ Results from MCMC estimates of the magnetization parameter $\beta$ in the Ising model. 
  Vertical line shows the true value.
  }
\end{figure}

@@

@@

  \item See if counting in nonoverlapping windows fixes the peakiness problem with TASEP.
